---
title: "STA572:Time Series Analysis and Forecasting"
subtitle: "C2. Univariate Modeling Techniques"
format: 
  revealjs:
    chalkboard: true
    fontsize: 20pt
    css: mystyle.css
    slide-number: true
author: "Lecturer, Department of Statistics"
footer: "Asmui Rahim, STA572 , Oct 2025"
execute: 
  echo: true
  warning: false
---

# Introduction

1.  Single Exponential Smoothing
2.  Holt's Method
3.  Holt-Winter's Trend and Seasonality
4.  Decomposition Method
5.  Forecasting using Decomposition Method

## Chapter 2: Data Preparation

### Setup Library

```{r}
library(fpp2)
library(forecast)
```

If no package, need to install.packages("fpp2") and/or install.packages ("forecast").

### Data Preparation

1.  Use "CPI Malaysia.csv" data. For this example, make sure the data is in your project directory.

```{r}
cpi <- read.csv("data/CPI Malaysia.csv",check.names=F)
cpits <- ts(cpi$`Consumer price index (2010 = 100)`, start=1960,frequency=1)
# estimation
est_part <- head(cpits, 0.8*length(cpits))
# evaluation
eva_part <- tail(cpits,0.2*length(cpits))
```

2.  Use "electric.csv" data.

```{r}
prod <- read.csv("data/electric.csv")
electric<-data.frame(t=seq(1:108),electric=prod[1:108,3])
electricts <- ts(electric$electric, start=c(2015,1), frequency=12) 
est_elec <- head(electricts, 0.8*length(electricts))
eva_elec <- tail(electricts, 0.2*length(electricts))
```

## Single Exponential Smoothing using R

```{r}
SE_est <- ses(est_part)
summary(SE_est)
```

------------------------------------------------------------------------

```{r}
SE_eva <- ses(eva_part, alpha=0.999)
summary(SE_eva)
```

------------------------------------------------------------------------

```{r}
SE_all <- ses(cpits,alpha=0.999)
summary(SE_all)
```

------------------------------------------------------------------------

```{r}
autoplot(cpits)+ylab("CPI")+
  autolayer(forecast(ses(cpits,alpha=0.999),h=5))
```

------------------------------------------------------------------------

## Holt's Method using R

```{r}
Holt_est <- holt(est_part)
summary(Holt_est)
```

------------------------------------------------------------------------

```{r}
Holt_eva <- holt(eva_part, alpha=0.9922, 
                 beta=0.3708)
summary(Holt_eva)
```

------------------------------------------------------------------------

```{r}
autoplot(cpits)+
  ylab("CPI")+
  autolayer(forecast(holt(cpits, alpha=0.9922,
                          beta=0.3708), h=5))
```

------------------------------------------------------------------------

## Holt-Winter's Trend and Seasonality using R

```{r}
HW_est <- hw(est_elec)
summary(HW_est)
```

------------------------------------------------------------------------

```{r}
HW_eva <- hw(eva_elec, alpha=0.9253, beta=0.0004, gamma=0.0001)
summary(HW_eva)
```

------------------------------------------------------------------------

```{r}
autoplot(electricts)+
  xlab("Year")+
  ylab("Indeks")+
  labs(title = "Forecasting Electricity Index using Holt-Winters Method")+
  autolayer(forecast(hw(electricts, alpha=0.9253, beta=0.0004, gamma=0.0001), h=12))
```

------------------------------------------------------------------------

## Decomposition Method

```{r}

data("AirPassengers")
data <- AirPassengers

plot(data)
```

------------------------------------------------------------------------

### Plot decompose

```{r}
data_decompose <- decompose(data)
data_decompose_mul <- decompose(data, type="multiplicative")
plot(data_decompose)
```

------------------------------------------------------------------------

```{r}
plot(data_decompose_mul)
```

------------------------------------------------------------------------

### Trend adjusted time series (without trend component)

```{r}

# trend adjusted time series (additive model)
trend_adjusted_add <- data-data_decompose$trend
plot(trend_adjusted_add)
```

------------------------------------------------------------------------

```{r}
# trend adjusted time series (multiplicative model)
trend_adjusted_mul <- data / data_decompose_mul$trend
plot(trend_adjusted_mul)
```

------------------------------------------------------------------------

### Seasonal adjusted time series (without seasonal component)

**Step 1**: Adjusted the data by removing the seasonal component.

Example of additive assumption.

```{r}
#| fig-align: "center"

# seasonal adusted time series
season_adjusted_add = data - data_decompose$seasonal
plot(season_adjusted_add)
```

------------------------------------------------------------------------

Example of multiplicative assumption.

```{r}
season_adjusted_mul <- data / data_decompose_mul$seasonal
plot(season_adjusted_mul)
```

------------------------------------------------------------------------

**Step 2**: Estimate the trend from seasonal adjusted data using linear trend model.

```{r}

time <- 1:length(season_adjusted_mul)

reg_deaseason <- lm(season_adjusted_mul~time)
summary(reg_deaseason)
```

------------------------------------------------------------------------

**Step 3**: Determine the seasonal components of the series.

```{r}

h <- 12
h_time <- (length(season_adjusted_mul)+1):(length(season_adjusted_mul)+h)
length(season_adjusted_mul)
print(h_time)
```

Estimated Seasonal Indices

```{r}
est_si <- as.numeric(tail(data_decompose_mul$seasonal,12))
print(est_si)
```

**Step 4**: Use linear trend model to estimate the $\hat{T}_{t+h}$

```{r}
h_time <- data.frame(time=h_time)
h_pred <- predict(reg_deaseason, newdata=h_time)
h_pred
```

------------------------------------------------------------------------

**Step 5**: Forecasting using decomposition method assuming multiplicative effect

```{r}
#| fig-align: "center"

forecast_mul <- h_pred * est_si

plot(1:length(data), data, type = "l", col = "blue", lwd = 2,
     xlim = c(1, length(data) + h),ylim = range(c(data, forecast_mul)),
     xlab = "Time", ylab = "Value", main = "Time Series Data and Forecast")

# Add forecast
lines((length(data) + 1):(length(data) + h), 
      forecast_mul, col = "red", lwd = 2)

# Add vertical line at forecast start
abline(v = length(data), lty = 2, col = "gray")

# Add legend
legend("topleft", legend = c("Original Data", "Forecast"),col = c("blue", "red"), lwd = 2)
```

------------------------------------------------------------------------

### Forecasting using Seasonal and Trend Decomposition using Loess (STL)

Robust method for decomposing time series data, allowing you to separately analyze the seasonal, trend, and remainder components.

Loess refers to a technique for estimating nonlinear relationships, and the STL method is especially useful because:

1.  It can handle any type of seasonality.
2.  The seasonal component can change over time, which can be controlled by the user.
3.  The trend-cycle smoothness can also be user-defined.
4.  It can be made robust to outliers.

------------------------------------------------------------------------

```{r}
#| fig-align: "center"
fit <- stl(data, t.window = 12, s.window = "periodic", robust=TRUE)
fit.sa <- seasadj(fit)
plot(fit.sa)
```

::: callout-note
## Note:

The STL algorithm is inherently additive, but logging the series converts a multiplicative relationship to an additive one in the log scale. When you back-transform (take exponentials) after decomposing, you retrieve the multiplicative structure for analysis and visualization.
:::

------------------------------------------------------------------------

```{r}
autoplot(naive(fit.sa))+
  ylab("New orders index")+
  ggtitle("Naive forecast of seasonality adjusted data")
```

Naive forecast of the seasonally adjusted air passenger data. The forecast then are "reseasonalised" by adding the seasonal naive forecast of the seasonal component.

------------------------------------------------------------------------

Use forecast () function applied to the stl object. Need to specify the method being used on the seasonally adjusted data, and the function will do the reseasonalising for you.

```{r}
autoplot(forecast(fit, method="naive",h=12))+
  ylab("New orders index")
```

------------------------------------------------------------------------

Shortcut approach using stlf() function.

What this function do?

1.  Decompose the time series using STL.
2.  Forecast the seasonally adjusted series.
3.  Return the reaseasonalised forecasts.

```{r}
#| fig-align: "center"

fcast <- stlf(data,method="naive",h=12)
autoplot(fcast)+
  ylab("New orders index")
```

------------------------------------------------------------------------

### Further reading:

<https://otexts.com/fpp2/expsmooth.html>

<https://otexts.com/fpp2/forecasting-decomposition.html>

# End of Slides
