---
title: "STA572:Time Series Analysis and Forecasting"
subtitle: "C4. Stochastic Process"
format: 
  revealjs:
    chalkboard: true
    fontsize: 20pt
    css: mystyle.css
    slide-number: true
author: "Lecturer, Department of Statistics"
footer: "Asmui Rahim, STA572 , Oct 2025"
execute: 
  echo: true
  warning: false
---

## Stationary Process

### Random process

Generate random process;

Example for $\phi_0=2$ and $\phi_0=8$.

```{r, echo=FALSE}
set.seed(151)
e <- rnorm(100,0,1)

y1 <- 8+e
y2 <- 2+e
x <- seq(1:length(y1))

plot(x,y1,type="l", col="blue", ylim=c(-2,11), ylab="")
lines(x, y2, type="l", col="green")
abline(lm(y1~x), col="red")
abline(lm(y2~x), col="red")
legend("bottomleft", legend=c("mean=8","mean=2"),
       col=c("blue","green"), lty=1, cex=0.8)

```

### Autoregressive Model

AR(1) model

```{r}
set.seed(151)
yt1 <- arima.sim(model=list(order=c(1,0,0), ar=0.2), n=100)
yt2 <- arima.sim(model=list(order=c(1,0,0), ar=0.5), n=100)
yt3 <- arima.sim(model=list(order=c(1,0,0), ar=0.8), n=100)

plot(yt1, type="l", col="blue", ylim=c(min(yt3),max(yt3)), ylab="")
lines(yt2, col="green")
lines(yt3, col="red")
legend("bottomright", legend=c("phi=0.2", "phi=0.5", "phi=0.8"), col=c("blue", "green", "red"), lty=1, cex=0.8)
```

## Non stationary process

### Random process

Example of non-stationary series.

```{r}
set.seed(111)
yt <- cumsum(rnorm(100,0,1))
plot(yt,type="l",ylab="")
```

## Checking for Non-stationary

Non-stationary series can be determined by:

-   Time series plot.

-   Autocorrelation function (ACF)

-   Statistical test- Augmented Dicky-Fuller (ADF) test

------------------------------------------------------------------------

Time series plot

```{r, echo=FALSE}

library(tseries)
library(forecast)

# Example 1: Stationary process (White Noise)
set.seed(123)
stationary_series <- rnorm(200, mean = 0, sd = 1)

# Example 2: Non-stationary process (Random Walk)
random_walk <- cumsum(rnorm(200))

# Example 3: Trend + Seasonality
time <- 1:200
trend_seasonal <- 0.05 * time + 5 * sin(2 * pi * time / 12) + rnorm(200)

# Plot comparison
par(mfrow = c(3, 1))
plot(stationary_series, type = "l", main = "Stationary: White Noise",
     ylab = "Value", xlab = "Time")
abline(h = mean(stationary_series), col = "red", lty = 2)

plot(random_walk, type = "l", main = "Non-Stationary: Random Walk",
     ylab = "Value", xlab = "Time", col = "darkred")

plot(trend_seasonal, type = "l", main = "Non-Stationary: Trend + Seasonal",
     ylab = "Value", xlab = "Time", col = "darkgreen")
```

------------------------------------------------------------------------

```{r, echo=FALSE}
par(mfrow = c(3, 1))

# ACF of stationary series - drops quickly
acf(stationary_series, main = "ACF: Stationary (White Noise)")

# ACF of random walk - slow decay (sign of non-stationarity)
acf(random_walk, main = "ACF: Non-Stationary (Random Walk)")

# ACF of trend+seasonal - slow decay with pattern
acf(trend_seasonal, main = "ACF: Non-Stationary (Trend+Seasonal)")

# Interpretation:
# - White noise: ACF ≈ 0 for all lags > 0
# - Random walk: ACF slowly decays, stays high
# - Trend+Seasonal: ACF shows pattern and slow decay
```

------------------------------------------------------------------------

Augmented Dickey-Fuller Test

```{r, warning=FALSE}
library(tseries)

# Test 1: White Noise (expect stationary)
adf_test1 <- adf.test(stationary_series)
adf_test1

# Test 2: Random Walk (expect non-stationary)
adf_test2 <- adf.test(random_walk)
adf_test2

# Test 3: After differencing random walk
diff_random_walk <- diff(random_walk)
adf_test3 <- adf.test(diff_random_walk)
adf_test3
```

------------------------------------------------------------------------

## Differencing in R

```{r}
unrate <- read.csv("employment.csv")
unratets <- ts(unrate$u_rate, start=1982)
plot(unratets, ylab="unemployement rate", xlab="Year", xaxt="n")
```

------------------------------------------------------------------------

```{r}
ur<-ts(unrate$u_rate, start = 1982)
diff_ur<-diff(ur)
autoplot(diff_ur)
```
