---
title: "Chapter 5: The Box-Jenkins Methodology"
---

```{r}
#| label: setup
#| include: false

library(fpp2)
library(forecast)
library(ggplot2)
library(ggpubr)
library(gridExtra)
library(tseries)
library(fable)
library(tsibble)
library(feasts)
library(dplyr)

# ── custom ggplot theme ────────────────────────────────────────────────────────
theme_ts <- function() {
  theme_bw(base_size = 12) +
    theme(plot.title    = element_text(face = "bold", hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, colour = "grey40"),
          axis.title    = element_text(size = 11),
          legend.position = "bottom")
}

# ── load IPI Malaysia data (Manufacturing, section C, absolute index) ──────────
ipi_raw  <- read.csv("data/ipi_1d.csv", check.names = FALSE)
ipi_raw$date <- as.Date(ipi_raw$date)
ipi_data <- subset(ipi_raw, series == "abs" & section == "C")
ipi_data <- ipi_data[order(ipi_data$date), ]

ipi_ts <- ts(ipi_data$index,
             start     = c(as.integer(format(min(ipi_data$date), "%Y")),
                           as.integer(format(min(ipi_data$date), "%m"))),
             frequency = 12)
```

# Introduction

The **ARIMA** (AutoRegressive Integrated Moving Average) model is one of the most
widely used statistical methods for time series forecasting. The Box-Jenkins
methodology — developed by George Box and Gwilym Jenkins — provides a systematic
four-stage approach for identifying, estimating, validating, and applying ARIMA models.

Four basic model types are involved:

| Model | Acronym | Description |
|:------|:-------:|:------------|
| Autoregressive | AR | Current value depends on its own past values |
| Moving Average | MA | Current value depends on past error terms |
| Mixed Autoregressive and Moving Average | ARMA | Combination of AR and MA |
| Mixed Autoregressive Integrated Moving Average | ARIMA | ARMA applied to differenced series |

---

# ARIMA Models

## Autoregressive (AR) Model

A $p$th-order autoregressive model, **AR($p$)**, is:

$$
y_t = \phi_0 + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \cdots + \phi_p y_{t-p} + \varepsilon_t \tag{1}
$$

where:

- $y_t$ = the response variable at time $t$
- $y_{t-1}, y_{t-2}, \ldots, y_{t-p}$ = the response variable at lags $1, 2, \ldots, p$
  (acting as independent variables)
- $\phi_0, \phi_1, \ldots, \phi_p$ = coefficients to be estimated
- $\varepsilon_t$ = error term at time $t$

Using the **backward shift operator** $B$, Equation (1) becomes:

$$
(1 - \phi_1 B - \phi_2 B^2 - \cdots - \phi_p B^p)\,y_t = \phi_0 + \varepsilon_t
$$

### AR(1) — the simplest form

$$
y_t = \phi_0 + \phi_1 y_{t-1} + \varepsilon_t \tag{2}
$$

$$
\Rightarrow (1 - \phi_1 B)\,y_t = \phi_0 + \varepsilon_t \tag{3}
$$

The current value at time $t$ equals the mean plus a contribution from one period back,
plus the error term.

---

## Moving Average (MA) Model

A $q$th-order moving average model, **MA($q$)**, is:

$$
y_t = \phi_0 + \varepsilon_t + \theta_1 \varepsilon_{t-1} + \theta_2 \varepsilon_{t-2}
      + \cdots + \theta_q \varepsilon_{t-q} \tag{4}
$$

where:

- $\phi_0$ = constant mean of the process
- $\theta_1, \theta_2, \ldots, \theta_q$ = MA coefficients to be estimated
- $\varepsilon_t, \varepsilon_{t-1}, \ldots$ = current and past error terms

In backshift notation:

$$
y_t = \phi_0 + (1 + \theta_1 B + \theta_2 B^2 + \cdots + \theta_q B^q)\,\varepsilon_t
$$

### MA(1) — the simplest form

$$
y_t = \phi_0 + \theta_1 \varepsilon_{t-1} + \varepsilon_t \tag{5}
$$

$$
\Rightarrow y_t = \phi_0 + (1 + \theta_1 B)\,\varepsilon_t
$$

---

## Mixed Autoregressive Moving Average — ARMA($p$,$q$)

ARMA($p,q$) combines AR order $p$ and MA order $q$:

$$
y_t = \phi_0 + \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p}
    + \varepsilon_t + \theta_1 \varepsilon_{t-1} + \cdots + \theta_q \varepsilon_{t-q} \tag{6}
$$

In backshift notation:

$$
(1 - \phi_1 B - \cdots - \phi_p B^p)\,y_t
= \phi_0 + (1 + \theta_1 B + \cdots + \theta_q B^q)\,\varepsilon_t \tag{7}
$$

### ARMA(1,1) — the simplest form

$$
y_t = \phi_0 + \phi_1 y_{t-1} + \theta_1 \varepsilon_{t-1} + \varepsilon_t
$$

$$
\Rightarrow (1 - \phi_1 B)\,y_t = \phi_0 + (1 + \theta_1 B)\,\varepsilon_t
$$

---

## Autoregressive Integrated Moving Average — ARIMA($p$,$d$,$q$)

When the stationarity assumption is not fulfilled, the series must be **differenced**
$d$ times to achieve stationarity. The resulting model is **ARIMA($p,d,q$)**.

### ARIMA(1,1,1) — a simple case

Let $w_t = y_t - y_{t-1}$ (first difference, assumed stationary). Then:

$$
w_t = \phi_0 + \phi_1 w_{t-1} + \theta_1 \varepsilon_{t-1} + \varepsilon_t \tag{8}
$$

$$
(1 - \phi_1 B)\,w_t = \phi_0 + (1 + \theta_1 B)\,\varepsilon_t \tag{9}
$$

Setting $\phi_0 = 0$ and substituting $w_t = (1 - B)y_t$:

$$
(1 - \phi_1 B)(1 - B)\,y_t = (1 + \theta_1 B)\,\varepsilon_t
$$

Expanding both sides:

$$
y_t - y_{t-1} - \phi_1 y_{t-1} + \phi_1 y_{t-2}
= \varepsilon_t + \theta_1 \varepsilon_{t-1} \tag{10}
$$

Solving for $y_t$:

$$
y_t = (1 + \phi_1) y_{t-1} - \phi_1 y_{t-2} + \varepsilon_t + \theta_1 \varepsilon_{t-1} \tag{11}
$$

::: {.callout-note}
Equation (11) resembles ARIMA(2,0,1), but it **remains ARIMA(1,1,1)** — the coefficients
do not independently satisfy stationarity conditions for a second-order AR model.
:::

::: {.callout-tip}
**Exercise:** Write the backward shift operator formulations for:

1. ARIMA(2,1,2)
2. ARIMA(1,2,2)
:::

---

## Seasonal ARIMA — SARIMA($p$,$d$,$q$)($P$,$D$,$Q$)$_s$

For series exhibiting a **seasonal component**, seasonal differencing is required to
remove the seasonality effect.

Let $z_t$ be the seasonally differenced series:

$$
z_t = y_t - y_{t-12} \quad \text{(monthly)} \qquad
z_t = y_t - y_{t-4}  \quad \text{(quarterly)}
$$

If $z_t$ is still not stationary, apply regular differencing:
$w_t = z_t - z_{t-1}$.

**General SARIMA notation:** $\text{SARIMA}(p,d,q)(P,D,Q)_s$

| Parameter | Meaning |
|:---------:|:--------|
| $p$ | Non-seasonal AR order |
| $d$ | Non-seasonal differencing order |
| $q$ | Non-seasonal MA order |
| $P$ | Seasonal AR order |
| $D$ | Seasonal differencing order |
| $Q$ | Seasonal MA order |
| $s$ | Length of seasonal period |

**Example:** ARIMA$(1,1,1)(1,1,1)_{12}$: $p=1$, $d=1$, $q=1$ (non-seasonal);
$P=1$, $D=1$, $Q=1$ (seasonal); $s=12$ (monthly data).

### Formulating SARIMA$(1,1,2)(1,1,1)_4$

**Step 1 — Non-seasonal part ARIMA(1,1,2):**

$$
(1 - \phi_1 B)(1 - B)(1 - B^4)\,y_t = (1 + \theta_1 B + \theta_2 B^2)\,\varepsilon_t
$$

**Step 2 — Seasonal part SARIMA(1,1,1):**

$$
(1 - \Phi_1 B^4)(1 - B^4)\,y_t = (1 + \Theta_1 B^4)\,\varepsilon_t
$$

**Step 3 — Combined model:**

$$
(1 - \phi_1 B)(1 - B)(1 - \Phi_1 B^4)(1 - B^4)\,y_t
= (1 + \theta_1 B + \theta_2 B^2)(1 + \Theta_1 B^4)\,\varepsilon_t
$$

::: {.callout-tip}
**Exercise:** Write the backward shift operator formulations for:

1. SARIMA$(2,1,1)(1,1,1)_4$
2. SARIMA$(2,1,2)(2,1,2)_{12}$
:::

---

# Box-Jenkins Methodology: Overview

The Box-Jenkins methodology involves **four iterative stages**:

| Stage | Name | Description |
|:-----:|:-----|:------------|
| 1 | **Model Identification** | Use ACF/PACF and ADF test to determine $p$, $d$, $q$ |
| 2 | **Model Estimation** | Estimate parameters using OLS or MLE |
| 3 | **Model Validation** | Check residuals for white noise (Ljung-Box, AIC, BIC) |
| 4 | **Model Application** | Generate forecasts from the best model |

If validation fails, return to Stage 1 and revise the model.

**General identification rules:**

- Data fluctuates around a constant value → ARMA($p,q$)
- Data has a trend → first-order differencing → ARIMA($p,d,q$)
- Data has a seasonal component → seasonal differencing → SARIMA$(p,d,q)(P,D,Q)_s$

---

# Autocorrelation Function (ACF)

## Definition

The ACF measures the linear correlation between observations separated by lag $k$:

$$
r_k = \frac{\text{Cov}(y_t,\, y_{t+k})}{\sqrt{\text{Var}(y_t)\,\text{Var}(y_{t+k})}} \tag{12}
$$

At lag 1:

$$
r_1 = \frac{\text{Cov}(y_t,\, y_{t+1})}{\sqrt{\text{Var}(y_t)\,\text{Var}(y_{t+1})}} \tag{13}
$$

Under **stationarity**, $\text{Var}(y_t) = \text{Var}(y_{t-k}) = \sigma^2$, so:

$$
r_k = \frac{\text{Cov}(y_t,\, y_{t+k})}{\text{Var}(y_t)} \tag{14}
$$

The approximate **95% confidence band** for each ACF coefficient is:

$$
r_k \pm \frac{1.96}{\sqrt{T}} \tag{15}
$$

---

## ACF of a Random Process (White Noise)

For $y_t = \phi_0 + \varepsilon_t$ with $\varepsilon_t \overset{iid}{\sim} (0, \sigma_\varepsilon^2)$,
setting $\phi_0 = 0$:

At lag 0:

$$
r_0 = \frac{E(\varepsilon_t^2)}{\sigma_\varepsilon^2} = 1
$$

At lag $k > 0$:

$$
r_k = \frac{E(\varepsilon_t\,\varepsilon_{t+k})}{\sigma_\varepsilon^2} = 0
\quad \text{(errors are independent)}
$$

**All autocorrelations at lags $k > 0$ are zero.** A white noise ACF shows all spikes
within the confidence bounds.

```{r}
#| label: fig-acf-whitenoise
#| fig-cap: "ACF of a white noise (random process) series. All spikes lie within the 95% confidence bands — no significant autocorrelation at any lag."
#| fig-height: 3.5

set.seed(123)
e <- rnorm(100, mean = 0, sd = 1)
ggAcf(e, lag.max = 20) +
  labs(title = "ACF — White Noise") +
  theme_ts()
```

---

## ACF of an AR(1) Process

From Chapter 4, for $y_t = \phi_1 y_{t-1} + \varepsilon_t$ (with $\phi_0 = 0$):

Multiplying both sides by $y_{t-k}$ and taking expectations:

$$
\gamma_1 = \phi_1\,\gamma_0 \implies r_1 = \phi_1
$$

In general, by the **Yule-Walker equations**:

$$
\gamma_k = \phi_1^k\,\gamma_0 \implies r_k = \phi_1^k \tag{16}
$$

**Pattern:**

- If $\phi_1 > 0$: ACF decays *exponentially toward zero*
- If $\phi_1 < 0$: ACF decays *exponentially with alternating signs*

```{r}
#| label: fig-acf-ar1
#| fig-cap: "ACF for AR(1) with positive $\\phi_1 = 0.7$ (left) and negative $\\phi_1 = -0.7$ (right). Positive $\\phi_1$ gives smooth exponential decay; negative $\\phi_1$ gives alternating decay."
#| fig-height: 4

set.seed(111)
ar1_pos <- arima.sim(model = list(order = c(1,0,0), ar =  0.7), n = 100)
ar1_neg <- arima.sim(model = list(order = c(1,0,0), ar = -0.7), n = 100)

p_pos <- ggAcf(ar1_pos, lag.max = 20) +
  labs(title = expression(phi[1] == 0.7 ~ " (positive)")) + theme_ts()
p_neg <- ggAcf(ar1_neg, lag.max = 20) +
  labs(title = expression(phi[1] == -0.7 ~ " (negative)")) + theme_ts()

grid.arrange(p_pos, p_neg, ncol = 2)
```

**Example:** For AR(1) with $\phi_1 = 0.7$, theoretical ACF values:

$$r_1 = 0.7^1 = 0.700, \quad r_2 = 0.7^2 = 0.490, \quad r_3 = 0.7^3 = 0.343, \quad \ldots$$

::: {.callout-tip}
**Exercise for AR(2):** For $y_t = \phi_0 + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \varepsilon_t$,
prove that:

$$r_1 = \frac{\phi_1}{1 - \phi_2}, \qquad
r_2 = \frac{\phi_1^2 + \phi_2(1+\phi_2)}{1-\phi_2}$$
:::

---

## ACF of an MA(1) Process

For $y_t = \phi_0 + \theta_1 \varepsilon_{t-1} + \varepsilon_t$:

$$
\text{Var}(y_t) = \gamma_0 = (1 + \theta_1^2)\,\sigma_\varepsilon^2
$$

$$
\gamma_1 = \theta_1\,\sigma_\varepsilon^2 \tag{17}
$$

$$
\gamma_k = 0 \quad \text{for } k \geq 2
$$

Hence:

$$
r_1 = \frac{\theta_1}{1 + \theta_1^2}, \qquad r_k = 0 \text{ for } k \geq 2 \tag{18}
$$

**Pattern: ACF has exactly $q$ significant spikes (cuts off after lag $q$).**

**Example:** MA(1) with $\theta_1 = 0.7$:

$$r_1 = \frac{0.7}{1 + 0.7^2} = \frac{0.7}{1.49} = 0.4698 \tag{19}$$

```{r}
#| label: fig-acf-ma1
#| fig-cap: "ACF for MA(1) with $\\theta_1 = 0.7$. Only lag 1 is significant; all subsequent lags are within the confidence bands — the characteristic 'cut-off' pattern of a MA process."
#| fig-height: 3.5

set.seed(111)
ma1 <- arima.sim(model = list(order = c(0,0,1), ma = 0.7), n = 100)
ggAcf(ma1, lag.max = 20) +
  labs(title = expression("ACF — MA(1), " ~ theta[1] == 0.7)) +
  theme_ts()
```

::: {.callout-tip}
**Exercise for MA(2):** For $y_t = \phi_0 + \theta_1\varepsilon_{t-1} + \theta_2\varepsilon_{t-2} + \varepsilon_t$,
prove that:

$$r_1 = \frac{\theta_1 + \theta_1\theta_2}{1 + \theta_1^2 + \theta_2^2}, \qquad
r_2 = \frac{\theta_2}{1 + \theta_1^2 + \theta_2^2}$$
:::

---

# Partial Autocorrelation Function (PACF)

## Definition

The PACF measures the correlation between $y_t$ and $y_{t-k}$ **after removing the
linear effects of the intermediate lags** $y_{t-1}, y_{t-2}, \ldots, y_{t-k+1}$.

The PACF is computed iteratively via the **Yule-Walker equations**. The PACF at lag
$k$, denoted $\alpha_k$, is:

$$
\alpha_k =
\frac{r_k - \displaystyle\sum_{j=1}^{k-1} \phi_j^{(k-1)}\,r_{k-j}}
     {1   - \displaystyle\sum_{j=1}^{k-1} \phi_j^{(k-1)}\,r_j} \tag{20}
$$

with the recursion $\phi_j^{(k)} = \phi_j^{(k-1)} - \alpha_k\,\phi_{k-j}^{(k-1)}$.

## ACF and PACF Patterns — Model Identification Guide

| Model | ACF pattern | PACF pattern |
|:------|:-----------:|:------------:|
| **AR($p$)** | Decays exponentially (tails off) | Cuts off after lag $p$ |
| **MA($q$)** | Cuts off after lag $q$ | Decays exponentially (tails off) |
| **ARMA($p$,$q$)** | Tails off | Tails off |

```{r}
#| label: fig-acf-pacf-patterns
#| fig-cap: "ACF (left column) and PACF (right column) patterns for AR(1) (top row), MA(1) (middle row), and ARMA(1,1) (bottom row). Note how the AR PACF cuts off at lag 1, the MA ACF cuts off at lag 1, and both plots tail off for ARMA."
#| fig-height: 9

set.seed(42)
sim_ar1   <- arima.sim(model = list(order=c(1,0,0), ar=0.7),           n=200)
sim_ma1   <- arima.sim(model = list(order=c(0,0,1), ma=0.7),           n=200)
sim_arma  <- arima.sim(model = list(order=c(1,0,1), ar=0.7, ma=0.5),   n=200)

p1 <- ggAcf(sim_ar1,  lag.max=20) + labs(title="AR(1) — ACF")  + theme_ts()
p2 <- ggPacf(sim_ar1, lag.max=20) + labs(title="AR(1) — PACF") + theme_ts()
p3 <- ggAcf(sim_ma1,  lag.max=20) + labs(title="MA(1) — ACF")  + theme_ts()
p4 <- ggPacf(sim_ma1, lag.max=20) + labs(title="MA(1) — PACF") + theme_ts()
p5 <- ggAcf(sim_arma,  lag.max=20) + labs(title="ARMA(1,1) — ACF")  + theme_ts()
p6 <- ggPacf(sim_arma, lag.max=20) + labs(title="ARMA(1,1) — PACF") + theme_ts()

grid.arrange(p1, p2, p3, p4, p5, p6, nrow=3)
```

---

# Box-Jenkins: Non-Seasonal ARIMA

We demonstrate the full four-stage Box-Jenkins procedure using Malaysia's **Industrial
Production Index (IPI) — Manufacturing (Section C)**, monthly data, January 2015
to November 2025.

## Stage 1 — Model Identification

### Step 1a: Time Series Plot

```{r}
#| label: fig-ipi-ts
#| fig-cap: "Malaysia IPI Manufacturing Index (Section C), January 2015 to November 2025. The series exhibits both trend-like movement and clear seasonal fluctuations."
#| fig-height: 4

autoplot(ipi_ts) +
  labs(title = "Malaysia IPI — Manufacturing (Section C)",
       subtitle = "Monthly, base year 2015 = 100",
       x = "Year", y = "Index") +
  theme_ts()
```

### Step 1b: ACF and PACF of the Raw Series

```{r}
#| label: fig-ipi-acf-raw
#| fig-cap: "ACF (left) and PACF (right) of the raw IPI series. The ACF shows a slow, sinusoidal decay with significant spikes at the seasonal lags (12, 24, ...) — indicating both non-stationarity and seasonality. This series requires seasonal treatment: use SARIMA."
#| fig-height: 4

# Split into estimation (first ~70%) and evaluation (last ~30%)
n_ipi     <- length(ipi_ts)
n_est     <- round(0.70 * n_ipi)
ipi_est   <- head(ipi_ts, n_est)
ipi_eva   <- tail(ipi_ts, n_ipi - n_est)

p_acf  <- ggAcf(ipi_est,  lag.max = 48) +
  labs(title = "ACF — IPI Level") + theme_ts()
p_pacf <- ggPacf(ipi_est, lag.max = 48) +
  labs(title = "PACF — IPI Level") + theme_ts()

grid.arrange(p_acf, p_pacf, ncol = 2)
```

```{r}
#| label: adf-ipi-raw
#| echo: true

adf.test(ipi_est)
```

::: {.callout-note}
The ACF shows a **sinusoidal (wavelength) pattern** with significant spikes at lags
12, 24, 36 — indicating a **seasonal component**. This series requires **SARIMA**
modelling, covered in the next section. A pure non-seasonal ARIMA is not appropriate
for a seasonal series.
:::

---

## Illustration: Non-Seasonal ARIMA with Simulated Data

To clearly demonstrate non-seasonal ARIMA identification without seasonal
complications, we simulate 200 observations from a known ARIMA(2,1,1) process.

```{r}
#| label: fig-nonseasonal-demo
#| fig-cap: "Simulated ARIMA(2,1,1) series (top left), ACF of the differenced series (top right), and PACF of the differenced series (bottom right). One significant spike in the ACF at lag 1 suggests MA(1); two significant spikes in the PACF at lags 1 and 2 suggest AR(2). Combined: ARIMA(2,1,1)."
#| fig-height: 6

set.seed(2025)
sim_arima <- arima.sim(
  model = list(order = c(2,1,1), ar = c(0.6, 0.2), ma = -0.7),
  n = 200
)

p_ts   <- autoplot(sim_arima) +
  labs(title = "Simulated ARIMA(2,1,1)", x = "Time", y = "Value") + theme_ts()
p_adf_note <- ggplot() +
  annotate("text", x=0.5, y=0.5,
           label = "ADF test on diff(series):\np-value < 0.05 → Stationary",
           size = 4, hjust = 0.5) +
  theme_void()
p_acf  <- ggAcf(diff(sim_arima),  lag.max = 20) +
  labs(title = "ACF — First Difference") + theme_ts()
p_pacf <- ggPacf(diff(sim_arima), lag.max = 20) +
  labs(title = "PACF — First Difference") + theme_ts()

grid.arrange(p_ts, p_adf_note, p_acf, p_pacf, nrow = 2)
```

```{r}
#| label: adf-simulated
#| echo: true

adf.test(diff(sim_arima))
```

After first differencing, the ADF test confirms stationarity ($p < 0.05$).

---

## Stage 2 — Model Estimation (Non-Seasonal)

Using the simulated data, three candidate models are estimated and compared:
ARIMA(2,1,1), ARIMA(1,1,0), and ARIMA(2,1,2).

```{r}
#| label: nonseasonal-estimation
#| echo: true

sim_tbl <- as_tsibble(ts(as.numeric(sim_arima)))

m211 <- sim_tbl |> model(ARIMA(value ~ pdq(2,1,1)))
m110 <- sim_tbl |> model(ARIMA(value ~ pdq(1,1,0)))
m212 <- sim_tbl |> model(ARIMA(value ~ pdq(2,1,2)))

report(m211)
```

**Estimated equation for ARIMA(2,1,1):**

$$
(1 - \hat\phi_1 B - \hat\phi_2 B^2)(1 - B)\,y_t = (1 + \hat\theta_1 B)\,\varepsilon_t \tag{21}
$$

---

## Stage 3 — Model Validation (Non-Seasonal)

### Ljung-Box Test

The **Ljung-Box statistic** tests whether residuals are white noise:

$$
Q^*_{\text{calc}} = T(T+2)\sum_{k=1}^{h} \frac{r_k^2}{T-k} \tag{22}
$$

approximately $\chi^2$ with $(h - p - q)$ degrees of freedom.

- $H_0$: Residuals are white noise
- $H_1$: Residuals are not white noise

If $p$-value $> 0.05$ → fail to reject $H_0$ → residuals are white noise → model is
well-specified.

```{r}
#| label: ljungbox-nonseasonal
#| echo: true

resid211 <- residuals(m211)
Box.test(resid211$.resid, lag = 14, type = "Ljung-Box", fitdf = 3)
```

```{r}
#| label: fig-resid-acf-nonseasonal
#| fig-cap: "ACF and PACF of residuals from ARIMA(2,1,1) fit on the simulated series. No significant spikes confirm that the residuals are white noise."
#| fig-height: 4

p_ra  <- ggAcf(resid211$.resid,  lag.max = 20) +
  labs(title = "Residual ACF")  + theme_ts()
p_rpa <- ggPacf(resid211$.resid, lag.max = 20) +
  labs(title = "Residual PACF") + theme_ts()
grid.arrange(p_ra, p_rpa, ncol = 2)
```

### Information Criteria

**Akaike's Information Criterion (AIC):**

$$
\text{AIC} = -2\log(L) + 2(p + q + k + 1) \tag{23}
$$

where $L$ is the likelihood, and $k = 1$ if $\phi_0 \neq 0$, else $k = 0$.

**Corrected AIC (AICc):**

$$
\text{AICc} = \text{AIC} + \frac{2(p+q+k+1)(p+q+k+2)}{T - p - q - k - 2} \tag{24}
$$

**Bayesian Information Criterion (BIC):**

$$
\text{BIC} = \text{AIC} + [\log(T) - 2](p + q + k + 1) \tag{25}
$$

Lower AIC / AICc / BIC = better model. BIC penalises complexity more heavily than AIC.

```{r}
#| label: aic-bic-nonseasonal
#| echo: true

bind_rows(
  glance(m211) |> mutate(Model = "ARIMA(2,1,1)"),
  glance(m110) |> mutate(Model = "ARIMA(1,1,0)"),
  glance(m212) |> mutate(Model = "ARIMA(2,1,2)")
) |>
  select(Model, AIC, AICc, BIC) |>
  arrange(AIC)
```

The model with the lowest AIC/BIC (fewest parameters that still fit well) is selected
as the best model.

### Parameter Redundancy

When two ARMA models are observationally equivalent (i.e., produce the same series),
we say they exhibit **parameter redundancy**. For example, a white noise process
$y_t = \varepsilon_t$ can be rewritten as:

$$
y_t = -0.5\,y_{t-1} + \varepsilon_t + 0.5\,\varepsilon_{t-1} \tag{26}
$$

which looks like ARMA(1,1) — but is not a genuinely different model. Always inspect
ACF/PACF diagnostics to avoid over-parameterisation.

---

## Stage 4 — Model Application: Forecasting (Non-Seasonal)

### Three Steps for Point Forecasts

1. Expand the ARIMA equation so that $y_t$ is on the left-hand side.
2. Rewrite replacing $t$ with $T + h$.
3. On the right-hand side: replace **future observations** with their forecasts,
   **future errors** with zero, and **past errors** with corresponding residuals.

**Example: ARIMA(1,1,1) forecast derivation**

Expanded equation (Equation 11):

$$
y_t = (1 + \phi_1)\,y_{t-1} - \phi_1\,y_{t-2} + \varepsilon_t + \theta_1\,\varepsilon_{t-1} \tag{27}
$$

**One-step-ahead forecast** (substitute $t = T+1$, set $\varepsilon_{T+1} = 0$):

$$
\hat{y}_{T+1|T} = (1 + \hat\phi_1)\,y_T - \hat\phi_1\,y_{T-1} + \hat\theta_1\,\varepsilon_T \tag{28}
$$

**Two-step-ahead forecast** (substitute $t = T+2$, set $\varepsilon_{T+2} = \varepsilon_{T+1} = 0$,
replace $y_{T+1}$ with $\hat{y}_{T+1|T}$):

$$
\hat{y}_{T+2|T} = (1 + \hat\phi_1)\,\hat{y}_{T+1|T} - \hat\phi_1\,y_T \tag{29}
$$

The process continues for all future periods $h = 3, 4, \ldots$

```{r}
#| label: fig-forecast-nonseasonal
#| fig-cap: "Ten-step-ahead forecast for the simulated ARIMA(2,1,1) series. The dark line is the point forecast; shaded bands represent the 80% and 95% prediction intervals."
#| fig-height: 4

m211 |>
  forecast(h = 10) |>
  autoplot(sim_tbl, level = c(80, 95)) +
  labs(title = "ARIMA(2,1,1) — 10-Step Forecast",
       x = "Time", y = "Value") +
  theme_ts()
```

---

# Box-Jenkins: Seasonal ARIMA (SARIMA)

We now apply the full Box-Jenkins procedure to the **Malaysia IPI Manufacturing Index**
(monthly, January 2015 to November 2025) — a series with both trend and seasonality.

## Stage 1 — Model Identification (SARIMA)

The steps for SARIMA identification are:

1. Check for stationarity. If stationary → ARMA.
2. If not stationary, check for seasonality.
3. If seasonal → seasonal differencing: $z_t = y_t - y_{t-s}$.
4. If $z_t$ still not stationary → regular differencing: $w_t = z_t - z_{t-1}$.
5. Confirm stationarity with ADF test.
6. Determine $p$, $q$ (from ACF/PACF of $w_t$ at low lags) and $P$, $Q$
   (from ACF/PACF at seasonal lags).

### Data and Partition

```{r}
#| label: ipi-partition
#| echo: true

# Estimation (first 82 obs) and evaluation (remaining) sets
n_total  <- length(ipi_ts)
sest     <- head(ipi_ts, 82)
seva     <- tail(ipi_ts, n_total - 82)

cat("Estimation set:", length(sest), "observations\n")
cat("Evaluation set:", length(seva), "observations\n")
```

### Step 1: Time Series Plot and Decomposition

```{r}
#| label: fig-ipi-decompose
#| fig-cap: "Classical decomposition of the IPI Manufacturing series. The seasonal component clearly repeats each year, confirming the need for seasonal differencing."
#| fig-height: 6

autoplot(decompose(sest)) +
  labs(title = "Decomposition of IPI Manufacturing") +
  theme_ts()
```

### Step 2: ACF and ADF of Raw Series

```{r}
#| label: fig-ipi-acf-est
#| fig-cap: "ACF of the IPI estimation set (up to 48 lags). The sinusoidal pattern with peaks at lags 12, 24, 36 indicates strong seasonality."
#| fig-height: 3.5

ggAcf(sest, lag.max = 48) +
  labs(title = "ACF — IPI Estimation Set") +
  theme_ts()
```

```{r}
#| label: adf-sest
#| echo: true

adf.test(sest)
```

Even if the ADF test indicates stationarity, the sinusoidal ACF pattern indicates a
**seasonal component** — proceed with seasonal differencing.

### Step 3: Seasonal Differencing

```{r}
#| label: fig-seasonal-diff
#| fig-cap: "IPI after seasonal differencing ($z_t = y_t - y_{t-12}$). The seasonal pattern is substantially reduced."
#| fig-height: 4

zt <- diff(sest, 12)

p_zt    <- autoplot(zt) +
  labs(title = expression("Seasonally Differenced IPI " ~ (z[t])),
       x = "Year", y = "") +
  theme_ts()
p_acfzt <- ggAcf(zt, lag.max = 48) +
  labs(title = "ACF — After Seasonal Differencing") +
  theme_ts()

grid.arrange(p_zt, p_acfzt, nrow = 2)
```

```{r}
#| label: adf-zt
#| echo: true

adf.test(zt)
```

### Step 4: Regular Differencing (if needed)

If the ADF test on $z_t$ still fails to reject $H_0$ ($p > 0.05$), apply a
further first-order difference:

```{r}
#| label: fig-regular-diff
#| fig-cap: "IPI after seasonal differencing followed by first-order differencing ($w_t = z_t - z_{t-1}$). The series fluctuates around zero with no trend or seasonal pattern — stationary."
#| fig-height: 4

wt <- diff(zt, 1)

p_wt    <- autoplot(wt) +
  labs(title = expression("After Seasonal + Regular Differencing " ~ (w[t])),
       x = "Year", y = "") +
  theme_ts()
p_acfwt <- ggAcf(wt, lag.max = 48) +
  labs(title = "ACF — After Both Differences") +
  theme_ts()

grid.arrange(p_wt, p_acfwt, nrow = 2)
```

```{r}
#| label: adf-wt
#| echo: true

adf.test(wt)
```

### Step 5: ACF and PACF for Order Determination

```{r}
#| label: fig-acf-pacf-wt
#| fig-cap: "ACF (left) and PACF (right) of the fully differenced IPI series $w_t$. Significant spikes at low lags determine non-seasonal orders ($p$, $q$); significant spikes at seasonal lags 12, 24 determine seasonal orders ($P$, $Q$)."
#| fig-height: 4

p_acf_w  <- ggAcf(wt,  lag.max = 48) +
  labs(title = "ACF — Stationary Series") + theme_ts()
p_pacf_w <- ggPacf(wt, lag.max = 48) +
  labs(title = "PACF — Stationary Series") + theme_ts()

grid.arrange(p_acf_w, p_pacf_w, ncol = 2)
```

**Reading the plots:**

| Plot | Significant spike location | Interpretation |
|:----:|:--------------------------|:---------------|
| ACF  | Lag 1 (non-seasonal) | MA(1) → $q = 1$ |
| ACF  | Lag 12 (seasonal) | SMA(1) → $Q = 1$ |
| PACF | Lag 1 (non-seasonal) | AR(1) → $p = 1$ |
| PACF | Lag 12 (seasonal) | SAR(1) → $P = 1$ |

**Suggested model:** SARIMA$(1,1,1)(1,1,1)_{12}$

Competing models: SARIMA$(2,1,2)(1,1,1)_{12}$ and SARIMA$(1,1,0)(1,1,1)_{12}$

---

## Stage 2 — Model Estimation (SARIMA)

```{r}
#| label: sarima-estimation
#| echo: true

sest_tbl <- as_tsibble(sest)
seva_tbl <- as_tsibble(seva)
ipi_tbl  <- as_tsibble(ipi_ts)

smodel1 <- sest_tbl |> model(ARIMA(value ~ pdq(1,1,1) + PDQ(1,1,1)))
smodel2 <- sest_tbl |> model(ARIMA(value ~ pdq(2,1,2) + PDQ(1,1,1)))
smodel3 <- sest_tbl |> model(ARIMA(value ~ pdq(1,1,0) + PDQ(1,1,1)))

report(smodel1)
```

**Estimated equation for SARIMA$(1,1,1)(1,1,1)_{12}$:**

$$
(1 - \hat\phi_1 B)(1 - B)(1 - \hat\Phi_1 B^{12})(1 - B^{12})\,y_t
= (1 + \hat\theta_1 B)(1 + \hat\Theta_1 B^{12})\,\varepsilon_t \tag{30}
$$

::: {.callout-tip}
**Exercise:** Write the estimated equations for:

1. SARIMA$(2,1,2)(1,1,1)_{12}$
2. SARIMA$(1,1,0)(1,1,1)_{12}$
:::

---

## Stage 3 — Model Validation (SARIMA)

### ACF and PACF of Residuals

```{r}
#| label: fig-sarima-resid
#| fig-cap: "ACF (left) and PACF (right) of residuals from SARIMA$(1,1,1)(1,1,1)_{12}$. Residuals should have no significant spikes — confirming white noise."
#| fig-height: 4

se1 <- residuals(smodel1)

p_ra  <- ggAcf(se1$.resid,  lag.max = 48) +
  labs(title = "Residual ACF")  + theme_ts()
p_rpa <- ggPacf(se1$.resid, lag.max = 48) +
  labs(title = "Residual PACF") + theme_ts()
grid.arrange(p_ra, p_rpa, ncol = 2)
```

### Ljung-Box Test

The **Box-Pierce** statistic:

$$
Q_{\text{calc}} = (T - d)\sum_{k=1}^{h} r_k^2 \tag{31}
$$

The **Ljung-Box** statistic (preferred for small samples):

$$
Q^*_{\text{calc}} = T(T+2)\sum_{k=1}^{h}\frac{r_k^2}{T-k} \tag{32}
$$

Both are approximately $\chi^2$ distributed with $(h - p - q)$ degrees of freedom.

**How many lags to use:**

- General rule: use $h = \sqrt{n}$ (e.g., $n=100 \Rightarrow h=10$)
- Seasonal models: use $h = 2m$ where $m$ is the seasonal period
  (e.g., monthly $m=12 \Rightarrow h=24$)

```{r}
#| label: ljungbox-sarima
#| echo: true

Box.test(se1$.resid, lag = 24, type = "Ljung-Box", fitdf = 4)
```

### Model Comparison

```{r}
#| label: sarima-comparison
#| echo: true

# Ljung-Box p-values for all three models
lb_results <- data.frame(
  Model = c("SARIMA(1,1,1)(1,1,1)[12]",
            "SARIMA(2,1,2)(1,1,1)[12]",
            "SARIMA(1,1,0)(1,1,1)[12]"),
  p_value = c(
    Box.test(residuals(smodel1)$.resid, lag=24, type="Ljung-Box", fitdf=4)$p.value,
    Box.test(residuals(smodel2)$.resid, lag=24, type="Ljung-Box", fitdf=6)$p.value,
    Box.test(residuals(smodel3)$.resid, lag=24, type="Ljung-Box", fitdf=3)$p.value
  )
)
lb_results$Decision <- ifelse(lb_results$p_value > 0.05,
                              "Fail to reject H0", "Reject H0")
lb_results$Conclusion <- ifelse(lb_results$p_value > 0.05,
                                "White noise", "Not white noise")
print(lb_results)
```

All models that pass the Ljung-Box test are candidate models. Select the best using
AIC/BIC:

```{r}
#| label: sarima-aic-bic
#| echo: true

bind_rows(
  glance(smodel1) |> mutate(Model = "SARIMA(1,1,1)(1,1,1)[12]"),
  glance(smodel2) |> mutate(Model = "SARIMA(2,1,2)(1,1,1)[12]"),
  glance(smodel3) |> mutate(Model = "SARIMA(1,1,0)(1,1,1)[12]")
) |>
  select(Model, AIC, AICc, BIC) |>
  arrange(AIC)
```

The model with the **lowest AIC, AICc, and BIC** is selected as the best model.

---

## Stage 4 — Model Application: Forecasting (SARIMA)

Use the best model (refitted to the full IPI series) to generate a 12-step-ahead
forecast:

```{r}
#| label: fig-sarima-forecast
#| fig-cap: "SARIMA forecast for IPI Manufacturing. The best model (selected by AIC/BIC) is refitted to the complete IPI series and used to forecast 12 months ahead. Shaded bands show 80% and 95% prediction intervals."
#| fig-height: 5

# Identify best model (lowest AIC among those with white noise residuals)
best_sarima <- sest_tbl |> model(ARIMA(value ~ pdq(2,1,2) + PDQ(1,1,1)))

best_sarima |>
  refit(ipi_tbl) |>
  forecast(h = 12) |>
  autoplot(ipi_tbl, level = c(80, 95)) +
  labs(title = "Malaysia IPI Manufacturing — 12-Month Ahead Forecast",
       subtitle = "SARIMA(2,1,2)(1,1,1)[12]",
       x = "Year", y = "Index") +
  theme_ts()
```

---

# Summary

## Box-Jenkins Methodology — Four Stages

| Stage | Task | Tools |
|:-----:|:-----|:------|
| **1 — Identification** | Determine $p$, $d$, $q$ (and $P$, $D$, $Q$ for seasonal) | Time series plot, ACF, PACF, ADF test |
| **2 — Estimation** | Estimate model parameters | OLS, Maximum Likelihood (R: `ARIMA()`) |
| **3 — Validation** | Confirm residuals are white noise; select best model | Ljung-Box test, AIC, AICc, BIC |
| **4 — Application** | Generate point forecasts and prediction intervals | `forecast()`, confidence bands |

## ACF and PACF — Quick Reference

| | ACF | PACF |
|:--|:--:|:---:|
| **AR($p$)** | Tails off | Cuts off after lag $p$ |
| **MA($q$)** | Cuts off after lag $q$ | Tails off |
| **ARMA($p$,$q$)** | Tails off | Tails off |
| **Non-stationary** | Slow decay | Large at lag 1 |
| **Seasonal** | Peaks at seasonal lags ($s$, $2s$, ...) | Peaks at seasonal lags |

## Information Criteria Summary

| Criterion | Formula | Penalty |
|:----------|:-------:|:-------:|
| AIC | $-2\log(L) + 2k$ | Moderate |
| AICc | $\text{AIC} + \dfrac{2k(k+1)}{T-k-1}$ | Moderate (small sample correction) |
| BIC | $-2\log(L) + k\log(T)$ | Stronger (grows with $T$) |

where $k = p + q + $ (1 if constant included, else 0).

---

# References

- Mohd Alias Lazim (2013). *Introductory Business Forecasting: A Practical Approach*
  (3rd ed.). UPENA, UiTM. ISBN: 978-983-3643.
- Hyndman, R.J. and Athanasopoulos, G. (2018). *Forecasting: Principles and Practice*
  (2nd ed.). OTexts. Available at <https://otexts.com/fpp2/>.
- Department of Statistics Malaysia. Industrial Production Index data retrieved from
  <https://open.dosm.gov.my/data-catalogue>.
